---
title: "Model Data Preparation"
author: "MBOA Analysis Team"
date: last-modified
---

```{r}
#| label: setup
#| cache: false
#| output: false
#| include: false
#| freeze: false

source("utils/dev_functions.R")
source("utils/manage_packages.R")

# Load all functions from the R package
load_all()

```

## Introductie

Dit document bereidt gegevens voor ten behoeve van voorspellende modellering van studentuitval over verschillende onderwijsprogramma's en niveaus. Het voorbereidingsproces omvat:

1. Laden van de benodigde datasets
2. Filteren en voorverwerken van gegevens
3. Opsplitsen van gegevens in trainings- en testsets
4. Opslaan van voorbereide gegevens in een georganiseerde structuur

## Gegevens laden

Als eerste laden we de gecombineerde en verrijkte inschrijvingsgegevens uit onze datapijplijn.

```{r}
#| label: load-data

# Load the data
enrollments_combined_enriched <- readRDS(file.path(config::get("data_combined_dir"),
                                                 "enrollments_combined_enriched.rds"))

enrollments_combined_enriched <- enrollments_combined_enriched %>%
  # Create shortened names
  mutate(
    TEAM_naam_kort = case_when(
      str_detect(TEAM_naam, "Beveiliging/Toezicht en Veiligheid") ~ "Beveiliging",
      str_detect(TEAM_naam, "Kappersschool Amersfoort") ~ "Kapperschool",
      str_detect(TEAM_naam, "Commercie en Evenementen") ~ "Commercie/Events",
      str_detect(TEAM_naam, "Hotelschool Amersfoort") ~ "Hotelschool",
      str_detect(TEAM_naam, "School voor Schoonheidsspecialisten") ~ "Schoonheid spec.",
      str_detect(TEAM_naam, "School voor Toerisme & Management") ~ "Toerisme & Mgt",
      str_detect(TEAM_naam, "Sportacademie Amersfoort") ~ "Sportacademie",
      str_detect(TEAM_naam, "Financiële Beroepen") ~ "Fin. Beroepen",
      str_detect(TEAM_naam, "MEI") ~ "MEI (Techniek)",
      .default = TEAM_naam
    ),
    OPLEIDING_crebo = as.numeric(gsub("[^0-9]", "", OPLEIDING_code)),
    # Handle potential NA values in creating shortened names
    OPLEIDING_naam_kort = case_when(
      is.na(OPLEIDING_naam) ~ "Onbekend",
      is.na(OPLEIDING_crebo) ~ word(OPLEIDING_naam, 1),
      TRUE ~ paste0(word(OPLEIDING_naam, 1), " (", OPLEIDING_crebo, ")")
    ),
    OPLEIDING_bc_label_kort = case_when(
      is.na(OPLEIDING_bc_label) ~ "Onbekend",
      is.na(OPLEIDING_bc_code) ~ word(OPLEIDING_bc_label, 3),
      TRUE ~ paste0(word(OPLEIDING_bc_label, 3), " (", OPLEIDING_bc_code, ")")
    )
  )

# Function to categorize reasons for leaving
categorize_reasons <- function(reason) {
  case_when(
      # Persoonlijk, of tenminste niet instelling-gerelateerd (Personal reasons)
      reason %in% c(
          "Verhuizing/reistijd",
          "Financi\xeble situatie",
          "Niet verschenen",
          "Verwijderd",
          "Detentie"
      ) ~ "Persoonsgebonden buiten invloed MBOA",
      
      reason %in% c(
          "Persoonlijk functioneren",
          "Persoonlijke omstandigheden",
          "Leerproblemen/onvoldoende capaciteit",
          "Onvoldoende taal/rekenen"
          
      ) ~ "Persoonsgebonden binnen invloed MBOA",
      
      # Werk (Work-related)
      reason %in% c(
          "Geen (erkend) leerbedrijf",
          "Ontslag of POK be\xebindigd",
          "Liever (fulltime) werken",
          "Werk of stage past niet",
          "Heeft arbeidsovereenkomst"
      ) ~ "Arbeidsmarkt",
      
      # Organisatie (Organizational)
      reason %in% c(
          "Schoolorganisatie",
          "Sociale/fysieke veiligheid",
          "Huisvesting/voorzieningen",
          "Onderwijs/didactiek",
          "Onvoldoende begeleiding"
      ) ~ "Instellingsgeboden",
      
      # Marketing (Expectations vs reality)
      reason %in% c(
          "Belangstelling andere opleiding/beroep",
          "Verkeerd beroepsbeeld",
          "Verkeerd opleidingsbeeld"
      ) ~ "Studie en beroepskeuze",
      
      reason %in% c(
          "No Show", 
          "Ambtshalve"
      ) ~ "Ambsthalve & No Show",
      
      reason %in% c(
          "Keuzedeel toevoegen",
          "DOORstroom: Organisatorisch",
          "DOORstroom: Examenstudent",
          "DOORstroom: Profiel => Uitstroomcrebo",
          "DOORstroom:Verlenger/Vertrager"
      ) ~ "Administratief",
      
      # Succes (Successful completion/progression)
      reason %in% c(
          "Certificaat",
          "Diploma",
          "Tweede diploma (in kalenderjaar)",
          "Zonder diploma, wel succesvol"
      ) ~ "Succes",
      
      is.na(reason) ~ "Nog steeds ingeschreven",
      # Everything else goes to "Overig" (Other)
      #  "Houding medewerkers",
      # "Administratieve  fout"
      #     "Persoonsgebonden",
      # "Besluit Student zie documenten",
      # "Te veel nevenactiviteiten",
      # "Besluit Intake vlg Stamkaart",
      TRUE ~ "Overig"
  )
}
```

```{r}
# Add categorized reasons
enrollments_combined_enriched <- enrollments_combined_enriched |>
  mutate(
    VERBINTENIS_reden_uitschrijving_categorie = categorize_reasons(VERBINTENIS_reden_uitschrijving)
  )

# Add dropout indicators per student and BC code
enrollments_combined_enriched <- enrollments_combined_enriched |>
  # Sort by student ID and start date to get chronological order
  arrange(DEELNEMER_ID, VERBINTENIS_begindatum) |>
  # Group by student and BC code
  group_by(DEELNEMER_ID, OPLEIDING_bc_label_kort) |>
  mutate(
    # Basic counts and details
    aantal_inschrijvingen = n(),
    DEELNEMER_BC_begindatum_eerst = min(VERBINTENIS_begindatum, na.rm = TRUE),
    DEELNEMER_BC_einddatum_laatst = if_else(
      is.na(max(VERBINTENIS_einddatum)),
      max(as.Date(VERBINTENIS_geplande_einddatum, format = "%d-%m-%Y")),
      max(VERBINTENIS_einddatum)
    ),
    
    # Total duration in days across all enrollments with this BC code
    DEELNEMER_BC_inschrijvingsduur = as.numeric(difftime(DEELNEMER_BC_einddatum_laatst, 
                                              DEELNEMER_BC_begindatum_eerst, 
                                              units = "days")),
    
    # Latest exit reason
    laatste_uitstroomreden = last(VERBINTENIS_reden_uitschrijving),
    laatste_uitstroomreden_categorie = last(VERBINTENIS_reden_uitschrijving_categorie),
    
    # Success indicators
    DEELNEMER_BC_uitval = !(DEELNEMER_BC_inschrijvingsduur > 365 | laatste_uitstroomreden_categorie == "Succes")
  ) |>
  ungroup()

```

```{r}

# Calculate age and add additional fields
enrollments_combined_enriched <- enrollments_combined_enriched |>
  mutate(DEELNEMER_leeftijd = DEELNEMER_BC_begindatum_eerst - DEELNEMER_geboorte_jaarmaand, 
         DEELNEMER_leeftijd = as.numeric(DEELNEMER_leeftijd, units = "days") / 365.25) |>
     select(DEELNEMER_ID,
           COHORT_startjaar,
           DEELNEMER_leeftijd,
           DEELNEMER_BC_uitval,
           DEELNEMER_geslacht,
           DEELNEMER_BC_begindatum_eerst,
           TEAM_naam_kort,
           AANMELDING_begin_dagen_tot_start,
           AANMELDING_afgerond_dagen_tot_start,
           OPLEIDING_naam_kort,
           OPLEIDING_bc_label_kort,
           DEELNEMER_havo_vwo_is_gezakt,
           VERBINTENIS_niveau,
           AANMELDING_laatst_gewijzigd_datum,
           DEELNEMER_BC_begindatum_eerst,
           DEELNEMER_BC_einddatum_laatst,
           DEELNEMER_BC_inschrijvingsduur,
           matches("VERBINTENIS_waarneming_pct_ongeoorloofd_week"),
           matches("VERBINTENIS_waarneming_pct_geoorloofd_week"),
           DEELNEMER_vooropleiding_categorie,
           DEELNEMER_passend_niveau,
           DEELNEMER_plaatsing,
           DEELNEMER_havo_vwo_is_gezakt,
           start_kwalificatie
           )
```

```{r}
# Function to reorganize week variables by removing all-NA columns and renumbering
reorganize_week_variables <- function(data, group_vars = c("COHORT_startjaar", "OPLEIDING_naam_kort")) {
  # Check if grouping variables exist in the data
  missing_vars <- setdiff(group_vars, names(data))
  if (length(missing_vars) > 0) {
    warning("The following grouping variables are missing from the data: ", 
            paste(missing_vars, collapse = ", "), 
            ". Using available grouping variables only.")
    group_vars <- intersect(group_vars, names(data))
    if (length(group_vars) == 0) {
      message("No valid grouping variables found. Processing the entire dataset as one group.")
    }
  }
  
  # Get the unique groups
  if (length(group_vars) > 0) {
    # Create grouping expression
    groups_list <- data %>%
      group_by(across(all_of(group_vars))) %>%
      group_keys()
  } else {
    # Create a dummy group if no valid grouping variables
    groups_list <- data.frame(dummy_group = "all_data")
    group_vars <- "dummy_group"
  }
  
  # Identify week variables that need processing
  week_vars <- names(data)[grepl("_week_[0-9]+$", names(data))]
  
  # If no week variables found, return original data with a message
  if(length(week_vars) == 0) {
    message("No week variables found in the dataset.")
    return(data)
  }
  
  # Identify the prefix patterns (e.g., "VERBINTENIS_waarneming_pct_ongeoorloofd", "VERBINTENIS_waarneming_pct_geoorloofd")
  prefixes <- unique(gsub("_week_[0-9]+$", "", week_vars))
  
  # Create a list to store group-specific data
  group_data_list <- list()
  
  # Track the weeks removed for each group
  weeks_removed_by_group <- list()
  
  # Initialize an empty data frame to capture all of the renaming that happens
  all_renamings <- data.frame(group = character(), 
                             old_var = character(), 
                             new_var = character(), 
                             stringsAsFactors = FALSE)
  
  # Process each group separately
  for(i in 1:nrow(groups_list)) {
    # Extract group values and create a filter
    if ("dummy_group" %in% names(groups_list)) {
      # Process all data as one group
      group_data <- data
      group_name <- "All data"
    } else {
      # Extract the current group's values
      current_group <- groups_list[i, , drop = FALSE]
      
      # Create filter conditions for each grouping variable
      filter_conditions <- list()
      for (var in group_vars) {
        filter_conditions[[var]] <- current_group[[var]]
      }
      
      # Filter data for the current group
      group_data <- data %>%
        filter(!!!rlang::quos(!!!lapply(seq_along(group_vars), function(j) {
          var <- group_vars[j]
          value <- current_group[[var]]
          rlang::expr(!!rlang::sym(var) == !!value)
        })))
      
      # Create a descriptive name for this group
      group_values <- unlist(current_group)
      group_name <- paste(group_vars, group_values, sep = ": ", collapse = ", ")
    }
    
    # Skip if the group has no data
    if (nrow(group_data) == 0) {
      message("Group [", group_name, "] has no data. Skipping.")
      next
    }
    
    # For this group, track which weeks are removed across all prefixes
    weeks_removed <- c()
    
    # Process each prefix for this group
    for(prefix in prefixes) {
      # Get all variables for this prefix
      prefix_vars <- names(group_data)[grepl(paste0("^", prefix, "_week_[0-9]+$"), names(group_data))]
      
      # Identify which columns are all NA for this group
      na_columns <- prefix_vars[sapply(prefix_vars, function(col) all(is.na(group_data[[col]])))]
      
      # Extract week numbers from NA columns
      if(length(na_columns) > 0) {
        week_nums <- unique(as.numeric(gsub(paste0("^", prefix, "_week_"), "", na_columns)))
        weeks_removed <- unique(c(weeks_removed, week_nums))
      }
      
      # Remove NA columns from group data
      if(length(na_columns) > 0) {
        group_data <- group_data %>% select(-all_of(na_columns))
      }
      
      # Keep only the variables that are not all NA for this prefix
      remaining_vars <- setdiff(prefix_vars, na_columns)
      
      if(length(remaining_vars) > 0) {
        # Extract week numbers from variable names
        week_numbers <- as.numeric(gsub(paste0("^", prefix, "_week_"), "", remaining_vars))
        
        # Create a mapping from old to new week numbers
        old_to_new <- data.frame(
          old_var = remaining_vars,
          old_num = week_numbers,
          new_num = rank(week_numbers),
          stringsAsFactors = FALSE
        ) %>%
          mutate(
            new_var = paste0(prefix, "_week_", sprintf("%02d", new_num))
          )
        
        # If there's any renaming to do (old names differ from new names)
        if(any(old_to_new$old_var != old_to_new$new_var)) {
          # Create a named vector for renaming
          rename_mapping <- setNames(old_to_new$old_var, old_to_new$new_var)
          
          # Capture the renaming for this group
          group_renamings <- data.frame(
            group = rep(group_name, sum(old_to_new$old_var != old_to_new$new_var)),
            old_var = old_to_new$old_var[old_to_new$old_var != old_to_new$new_var],
            new_var = old_to_new$new_var[old_to_new$old_var != old_to_new$new_var],
            stringsAsFactors = FALSE
          )
          all_renamings <- rbind(all_renamings, group_renamings)
          
          # Rename the columns in this group's data
          group_data <- group_data %>%
            rename(!!!rename_mapping)
        }
      }
    }
    
    # Store the weeks that were removed for this group
    if(length(weeks_removed) > 0) {
      weeks_removed_by_group[[group_name]] <- sort(weeks_removed)
    }
    
    # Store the processed group data
    group_data_list[[i]] <- group_data
  }
  
  # Combine all group data back together
  if (length(group_data_list) > 0) {
    result_data <- bind_rows(group_data_list)
  } else {
    message("No groups were processed. Returning original data.")
    return(data)
  }
  
  # Output summary of weeks removed by group
  message("\n===== WEEKS REMOVED BY GROUP =====")
  all_weeks_removed <- sort(unique(unlist(weeks_removed_by_group)))
  
  # First report weeks removed from all groups
  if(length(all_weeks_removed) > 0) {
    all_groups_weeks <- all_weeks_removed[sapply(all_weeks_removed, function(week) {
      all(sapply(weeks_removed_by_group, function(group_weeks) week %in% group_weeks))
    })]
    
    if(length(all_groups_weeks) > 0) {
      message("Weeks removed from ALL groups: ", 
              paste(sprintf("%02d", all_groups_weeks), collapse = ", "))
    }
    
    # Then report group-specific removals
    for(group in names(weeks_removed_by_group)) {
      group_specific <- setdiff(weeks_removed_by_group[[group]], all_groups_weeks)
      if(length(group_specific) > 0) {
        message("Weeks removed from group [", group, "]: ", 
                paste(sprintf("%02d", group_specific), collapse = ", "))
      }
    }
  } else {
    message("No weeks were removed from any group.")
  }
  
  # Output summary of renamings if any occurred
  if(nrow(all_renamings) > 0) {
    message("\n===== COLUMN RENAMINGS =====")
    
    # Group renamings by pattern for cleaner output
    renaming_patterns <- all_renamings %>%
      group_by(group) %>%
      summarise(
        old_patterns = paste(gsub("^.*week_", "week_", old_var), collapse = ", "),
        new_patterns = paste(gsub("^.*week_", "week_", new_var), collapse = ", ")
      )
    
    for(i in 1:nrow(renaming_patterns)) {
      message("Group [", renaming_patterns$group[i], "]:")
      message("  ", renaming_patterns$old_patterns[i], " -> ", renaming_patterns$new_patterns[i])
    }
  } else {
    message("\nNo column renamings were needed.")
  }
  
  return(result_data)
}

# Reorganize the week variables in the dataset
enrollments_combined_enriched <- reorganize_week_variables(enrollments_combined_enriched)

# Show basic data dimensions
cat("Loaded dataset dimensions:", dim(enrollments_combined_enriched), "\n")
```

## Identificeren van programma's en niveaus

We identificeren de beschikbare opleidingsprogramma's en onderwijsniveaus voor modellering.

```{r}
#| label: identify-programs-levels

# Get model settings from config
model_settings <- config::get("model_settings")
test_run <- model_settings$test_run
min_samples <- model_settings$min_training_samples
test_cohort_year <- model_settings$test_cohort_year

# Get unique programs and their counts, filtering for model-eligible cohort years
programs_level_count <- enrollments_combined_enriched |>
  filter(COHORT_startjaar %in% c(model_settings$training_cohort_years, test_cohort_year)) |>
  count(OPLEIDING_naam_kort, VERBINTENIS_niveau, COHORT_startjaar) |>
  pivot_wider(
    names_from = COHORT_startjaar,
    values_from = n,
    values_fill = 0
  ) |>
  rowwise() |>
  mutate(
    total_training = sum(c_across(as.character(model_settings$training_cohort_years))),
    total_test = !!as.name(as.character(test_cohort_year)),
    total = total_training + total_test
  ) |>
  ungroup() |>
  arrange(desc(total))

# Display available programs and their counts
cat("Available programs and levels by cohort year:\n")
print(programs_level_count |> arrange(desc(total)))

# Select programs based on config settings
if (test_run) {
  cat("\nRunning in TEST mode with limited programs and levels\n")
  
  # Use only the configured test programs and levels
  selected_programs <- model_settings$test_programs
  selected_levels <- model_settings$test_levels
  
  cat("Selected test programs:", paste(selected_programs, collapse=", "), "\n")
  cat("Selected test levels:", paste(selected_levels, collapse=", "), "\n")
} else {
  # In full mode, select programs with sufficient data
  filtered_programs <- programs_level_count |>
    filter(
      total_training >= model_settings$min_training_samples,
      total_test >= model_settings$min_test_samples
    )
  
  selected_programs <- unique(filtered_programs$OPLEIDING_naam_kort)
  selected_levels <- unique(filtered_programs$VERBINTENIS_niveau)
  
  cat("\nSelected", length(selected_programs), "programs with sufficient data:\n")
  print(filtered_programs |> select(OPLEIDING_naam_kort, VERBINTENIS_niveau, total_training, total_test, total))
}

# Save selected programs and levels for use in other QMD files
saveRDS(
  list(
    programs = selected_programs,
    levels = selected_levels,
    test_run = test_run,
    test_cohort_year = test_cohort_year
  ),
  file.path(config::get("modelled_dir"), "selected_programs_levels.rds")
)
```

## Gegevensvoorbereiding voor alle programma's

We bereiden eerst een dataset voor die alle programma's en niveaus bevat.

```{r}
#| label: prepare-all-data

# Define the week variable strategies to test
week_strategies <- c("none", "early", "all")

# Prepare data for all programs and levels with different week variable strategies
all_data_preps <- list()

for(week_strategy in week_strategies) {
  cat("\n--- Preparing all programs data with week strategy:", week_strategy, "---\n")
  
  all_data_preps[[week_strategy]] <- prepare_model_data(
    data = enrollments_combined_enriched,
    program_filter = NULL,
    level_filter = NULL,
    test_cohort = test_cohort_year,
    week_vars = week_strategy,
    save = TRUE
  )
  
  # Display information about the prepared data
  cat("All programs (", week_strategy, "weeks) - Training data dimensions:", 
      dim(all_data_preps[[week_strategy]]$train), "\n")
  cat("All programs (", week_strategy, "weeks) - Test data dimensions:", 
      dim(all_data_preps[[week_strategy]]$test), "\n")
  
  # Display variable counts for different strategies
  cat("Number of variables:", ncol(all_data_preps[[week_strategy]]$train), "\n")
  
  # Count week variables
  week_var_count <- sum(grepl("week", names(all_data_preps[[week_strategy]]$train)))
  cat("Number of week variables:", week_var_count, "\n")
  
  # Check class balance in training data
  train_balance <- table(all_data_preps[[week_strategy]]$train$DEELNEMER_BC_uitval)
  cat("Class balance in training data:\n")
  print(prop.table(train_balance))
}

# Show a comparison of variable counts across strategies
var_counts <- data.frame(
  strategy = week_strategies,
  total_vars = sapply(all_data_preps, function(x) ncol(x$train)),
  week_vars = sapply(all_data_preps, function(x) sum(grepl("week", names(x$train))))
)

cat("\nVariable count comparison across week strategies:\n")
print(var_counts)
```

## Gegevensvoorbereiding per programma en niveau

Nu gaan we afzonderlijke datasets voorbereiden voor elke geselecteerde programma-niveaucombinatie.

```{r}
#| label: prepare-by-program-level

# TODO: All combinations of programs - levels are checked and tried, while we now that many 
# combinations don't exist at all.

# Initialize a list to store all prepared datasets
program_level_data <- list()

# Prepare data for each program-level combination with different week variable strategies
for (program in selected_programs) {
  for (level in selected_levels) {
    # Create base identifier for this combination
    base_combo_id <- paste0(program, "_level", level)
    
    cat("\n===== Preparing data for:", base_combo_id, "=====\n")
    
    # Apply each week variable strategy
    for (week_strategy in week_strategies) {
      # Create full identifier including week strategy
      combo_id <- paste0(base_combo_id, "_weeks_", week_strategy)
      cat("\n--- Week strategy:", week_strategy, "---\n")
      
      # Try to prepare data for this combination
      tryCatch({
        program_level_data[[combo_id]] <- prepare_model_data(
          data = enrollments_combined_enriched,
          program_filter = program,
          level_filter = level,
          test_cohort = test_cohort_year,
          week_vars = week_strategy,
          save = TRUE
        )
        
        # Display information about the prepared data
        cat("  Training data dimensions:", dim(program_level_data[[combo_id]]$train), "\n")
        cat("  Test data dimensions:", dim(program_level_data[[combo_id]]$test), "\n")
        
        # Count week variables
        week_var_count <- sum(grepl("week", names(program_level_data[[combo_id]]$train)))
        cat("  Number of week variables:", week_var_count, "\n")
        
        # Check class balance
        if (nrow(program_level_data[[combo_id]]$train) > 0) {
          train_balance <- table(program_level_data[[combo_id]]$train$DEELNEMER_BC_uitval)
          cat("  Class balance in training data:", paste(names(train_balance), "=", train_balance, collapse=", "), "\n")
          
          # Check if it meets minimum requirements
          min_class <- min(train_balance)
          if (min_class < model_settings$min_minority_class) {
            cat("  WARNING: Minority class has only", min_class, "samples (minimum needed:", 
                model_settings$min_minority_class, ")\n")
          }
        }
        
      }, error = function(e) {
        cat("  Error preparing data:", e$message, "\n")
      })
    }
  }
}

# Count how many valid datasets we created per week strategy
valid_datasets_by_strategy <- numeric(length(week_strategies))
names(valid_datasets_by_strategy) <- week_strategies

for (i in seq_along(week_strategies)) {
  strategy <- week_strategies[i]
  pattern <- paste0("_weeks_", strategy, "$")
  relevant_keys <- grep(pattern, names(program_level_data), value = TRUE)
  
  # Count how many datasets are valid for this strategy
  valid_count <- 0
  if (length(relevant_keys) > 0) {
    for (key in relevant_keys) {
      x <- program_level_data[[key]]
      if (!is.null(x) && nrow(x$train) > 0 && nrow(x$test) > 0) {
        valid_count <- valid_count + 1
      }
    }
  }
  
  valid_datasets_by_strategy[i] <- valid_count
}

cat("\n===== Summary of prepared datasets =====\n")
for (i in seq_along(week_strategies)) {
  cat("Week strategy '", week_strategies[i], "': ", valid_datasets_by_strategy[i], 
      " valid program-level datasets\n", sep="")
}

total_valid <- sum(valid_datasets_by_strategy)
cat("\nSuccessfully prepared", total_valid, "valid program-level datasets in total\n")
```

## Dataset lijst opslaan voor de volgende stappen

```{r}
#| label: save-dataset-list

# Create a summary of all prepared datasets
dataset_summary <- data.frame(
  id = names(program_level_data),
  program_level = sapply(strsplit(names(program_level_data), "_weeks_"), function(x) x[1]),
  week_strategy = sapply(strsplit(names(program_level_data), "_weeks_"), function(x) x[2]),
  train_rows = sapply(program_level_data, function(x) if(!is.null(x)) nrow(x$train) else 0),
  test_rows = sapply(program_level_data, function(x) if(!is.null(x)) nrow(x$test) else 0),
  train_columns = sapply(program_level_data, function(x) if(!is.null(x)) ncol(x$train) else 0),
  week_vars = sapply(program_level_data, function(x) {
    if(!is.null(x)) sum(grepl("week", names(x$train))) else 0
  }),
  ratio_0 = sapply(program_level_data, function(x) {
    if(!is.null(x) && nrow(x$train) > 0) {
      table_values <- table(x$train$DEELNEMER_BC_uitval)
      names_values <- names(table_values)
      if(length(names_values) >= 2) {
        table_values[names_values[1]] / sum(table_values)
      } else {
        NA
      }
    } else {
      NA
    }
  }),
  ratio_1 = sapply(program_level_data, function(x) {
    if(!is.null(x) && nrow(x$train) > 0) {
      table_values <- table(x$train$DEELNEMER_BC_uitval)
      names_values <- names(table_values)
      if(length(names_values) >= 2) {
        table_values[names_values[2]] / sum(table_values)
      } else {
        NA
      }
    } else {
      NA
    }
  })
)

# Add min_class column
dataset_summary$min_class_count <- pmin(
  dataset_summary$ratio_0 * dataset_summary$train_rows,
  dataset_summary$ratio_1 * dataset_summary$train_rows,
  na.rm = TRUE
)

# Filter for valid datasets (meeting minimum requirements)
valid_datasets <- dataset_summary |>
  filter(
    train_rows >= model_settings$min_training_samples,
    test_rows >= model_settings$min_test_samples,
    min_class_count >= model_settings$min_minority_class
  )

# Save the summary
saveRDS(dataset_summary, file.path(config::get("modelled_dir"), "prepared", "dataset_summary.rds"))
saveRDS(valid_datasets, file.path(config::get("modelled_dir"), "prepared", "valid_datasets.rds"))

# Show the summary by week strategy
dataset_summary_by_strategy <- dataset_summary |>
  group_by(week_strategy) |>
  summarize(
    total_datasets = n(),
    valid_datasets = sum(train_rows >= model_settings$min_training_samples & 
                        test_rows >= model_settings$min_test_samples &
                        min_class_count >= model_settings$min_minority_class),
    avg_train_rows = mean(train_rows),
    avg_columns = mean(train_columns),
    avg_week_vars = mean(week_vars)
  )

cat("\nDataset summary by week strategy:\n")
print(knitr::kable(dataset_summary_by_strategy, digits = 1))

# Show detailed table of valid datasets
cat("\nValid datasets meeting minimum requirements:\n")
valid_summary <- valid_datasets |>
  select(id, program_level, week_strategy, train_rows, test_rows, week_vars, min_class_count) |>
  arrange(program_level, week_strategy)

print(knitr::kable(valid_summary, digits = 0))
```

## Volgende stappen

De voorbereide datasets zijn nu beschikbaar in de map `data/modelled/prepared` en klaar voor modeltraining. Het volgende document in deze workflow zal voorspellende modellen trainen met behulp van deze voorbereide datasets.
