---
title: "Model Training and Evaluation"
author: "MBOA Analysis Team"
date: last-modified
---

```{r}
#| label: setup
#| cache: false
#| output: false
#| include: false
#| freeze: false

source("utils/dev_functions.R")
source("utils/manage_packages.R")

# Load all functions from the R package
load_all()

```

## Introductie

Dit document traint voorspellende modellen voor studentuitval over verschillende onderwijsprogramma's en niveaus. Het proces omvat:

1. Laden van voorbereide datasets uit de vorige stap
2. Trainen van modellen voor elke programma-niveaucombinatie
3. Evalueren van modelprestaties
4. Opslaan van getrainde modellen en metrieken

## Laden van voorbereide gegevens

We laden eerst de samenvatting van de voorbereide datasets die in het vorige document zijn gemaakt.

```{r}
#| label: load-dataset-info

# Get model settings from config
model_settings <- config::get("model_settings")

# Load information about selected programs and levels
selection_path <- file.path(config::get("modelled_dir"), "selected_programs_levels.rds")
if(file.exists(selection_path)) {
  selection_info <- readRDS(selection_path)
  test_run <- selection_info$test_run
  test_cohort_year <- selection_info$test_cohort_year
  cat("Test run:", test_run, "\n")
  cat("Test cohort year:", test_cohort_year, "\n")
} else {
  test_run <- model_settings$test_run
  test_cohort_year <- model_settings$test_cohort_year
  cat("Warning: Selected programs file not found. Using config settings.\n")
}

# Load valid datasets
valid_datasets_path <- file.path(config::get("modelled_dir"), "prepared", "valid_datasets.rds")

if(file.exists(valid_datasets_path)) {
  valid_datasets_df <- readRDS(valid_datasets_path)
  
  # Show dataset summary by week strategy
  dataset_by_strategy <- valid_datasets_df |>
    group_by(week_strategy) |>
    summarize(
      count = n(),
      avg_train_rows = mean(train_rows),
      avg_week_vars = mean(week_vars),
      unique_programs = n_distinct(program_level)
    )
  
  cat("Found", nrow(valid_datasets_df), "valid datasets across", 
      n_distinct(valid_datasets_df$program_level), "program-levels\n\n")
  
  print(knitr::kable(dataset_by_strategy, digits = 1))
  
  # Extract dataset IDs for modeling
  valid_datasets <- valid_datasets_df$id
} else {
  # If valid datasets file not found, try the summary file
  dataset_summary_path <- file.path(config::get("modelled_dir"), "prepared", "dataset_summary.rds")
  
  if(file.exists(dataset_summary_path)) {
    dataset_summary <- readRDS(dataset_summary_path)
    
    # Filter for valid datasets using config settings
    valid_datasets_df <- dataset_summary |>
      filter(
        train_rows >= model_settings$min_training_samples,
        test_rows >= model_settings$min_test_samples, 
        min_class_count >= model_settings$min_minority_class
      )
    
    valid_datasets <- valid_datasets_df$id
    
    cat("Found", nrow(valid_datasets_df), "valid datasets\n")
  } else {
    # If neither file is found, look for RDS files directly
    prepared_files <- list.files(
      file.path(config::get("modelled_dir"), "prepared"), 
      pattern = "\\.rds$", 
      full.names = FALSE
    )
    
    # Filter out the summary file itself
    prepared_files <- prepared_files[!prepared_files %in% c("dataset_summary.rds", "valid_datasets.rds")]
    
    cat("Warning: Dataset summary files not found. Found", length(prepared_files), "prepared dataset files directly\n")
    valid_datasets <- gsub("\\.rds$", "", prepared_files)
  }
}

# Define the week strategies we're testing
week_strategies <- c("none", "early", "all")
```

## Modeltrainingsfunctie

We definiëren een hulpfunctie om modellen te trainen voor elke dataset.

```{r}
#| label: define-training-function

train_model_for_dataset <- function(dataset_id, model_type = "random_forest") {
  cat("\nTraining model for:", dataset_id, "\n")
  
  # Load prepared data
  data_path <- file.path(config::get("modelled_dir"), "prepared", paste0(dataset_id, ".rds"))
  
  if(!file.exists(data_path)) {
    # Try with cohort suffix if not found
    data_path <- file.path(config::get("modelled_dir"), "prepared", paste0(dataset_id, "_cohort", test_cohort_year, ".rds"))
    
    if(!file.exists(data_path)) {
      cat("  ERROR: Data file not found:", data_path, "\n")
      return(NULL)
    }
  }
  
  # Load the data
  prepared_data <- readRDS(data_path)
  
  # Check if data has sufficient rows
  ## TODO again hard-coded values
  if(nrow(prepared_data$train) < 50 || nrow(prepared_data$test) < 20) {
    cat("  SKIPPING: Insufficient data (train:", nrow(prepared_data$train), 
        ", test:", nrow(prepared_data$test), ")\n")
    return(NULL)
  }
  
  # Check class balance
  class_counts <- table(prepared_data$train$DEELNEMER_BC_uitval)
  min_class_count <- min(class_counts)
  
  ## TODO hard-coded values, should be in config
  if(min_class_count < 5) {
    cat("  SKIPPING: Insufficient samples in minority class (", 
        names(class_counts)[which.min(class_counts)], ":", min_class_count, ")\n")
    return(NULL)
  }
  
  # Run the model
  tryCatch({
    model_results <- run_model(
      data_list = prepared_data,
      model_type = model_type,
      grid_size = 20,
      n_folds = 5,
      save = TRUE
    )
    
    cat("  SUCCESS: Model trained successfully\n")
    
    # Return results
    return(model_results)
    
  }, error = function(e) {
    cat("  ERROR: Model training failed:", e$message, "\n")
    return(NULL)
  })
}
```

## Training van individuele programma-niveau modellen

Nu gaan we modellen trainen voor elke individuele programma-niveaucombinatie.

```{r}
#| label: train-program-level-models

# Initialize results list
model_results_list <- list()

# If test run, limit to a small subset of datasets
if(test_run) {
  # For test runs, limit to one program-level with all week strategies
  # Extract one program-level to test all week strategies
  unique_program_levels <- unique(str_remove(valid_datasets, "_weeks_.*$"))
  
  if(length(unique_program_levels) > 0) {
    test_program_level <- unique_program_levels[1]
    test_datasets <- valid_datasets[str_detect(valid_datasets, 
                                               stringr::fixed(paste0(test_program_level, "_weeks_")))]
    
    cat("Running in TEST mode with program-level:", test_program_level, "\n")
    cat("Testing", length(test_datasets), "week strategies for this program-level\n")
    
    datasets_to_train <- test_datasets
  } else {
    # Fallback if there are no valid datasets
    cat("No valid program-levels found. Using first 3 valid datasets for testing.\n")
    datasets_to_train <- head(valid_datasets, 3)
  }
} else {
  # In full mode, use all valid datasets
  datasets_to_train <- valid_datasets
}

# Train models for each selected dataset
for(dataset_id in datasets_to_train) {
  # Extract program-level and week strategy for better organization
  parts <- strsplit(dataset_id, "_weeks_")[[1]]
  program_level <- parts[1]
  week_strategy <- parts[2]
  
  cat("\n----- Training model for program-level:", program_level, "| Week strategy:", week_strategy, "-----\n")
  
  # Train model for this dataset
  model_result <- train_model_for_dataset(dataset_id)
  
  # Store result if successful
  if(!is.null(model_result)) {
    model_results_list[[dataset_id]] <- model_result
  }
}

# Summarize results by week strategy
cat("\n----- Training Summary -----\n")
successful_models <- names(model_results_list)
cat("Successfully trained", length(successful_models), "models\n\n")

# Count models by week strategy
for(strategy in week_strategies) {
  strategy_models <- successful_models[str_detect(successful_models, 
                                                  stringr::fixed(paste0("_weeks_", strategy)))]
  cat("Week strategy '", strategy, "': ", length(strategy_models), " models\n", sep="")
}

# Save the successfully trained model list
saveRDS(model_results_list, file.path(config::get("modelled_dir"), "run", "all_model_results.rds"))
```

## Samenvatting van modelprestaties

Laten we een samenvatting maken van de modelprestaties voor alle getrainde modellen.

```{r}
#| label: model-performance-summary

# Create a performance summary dataframe
performance_summary <- data.frame()

# Add individual model metrics with program-level and week strategy information
for(dataset_id in names(model_results_list)) {
  if(!is.null(model_results_list[[dataset_id]]$metrics)) {
    # Extract program-level and week strategy
    parts <- strsplit(dataset_id, "_weeks_")[[1]]
    program_level <- parts[1]
    week_strategy <- parts[2]
    
    metrics <- model_results_list[[dataset_id]]$metrics |>
      mutate(
        dataset_id = dataset_id,
        program_level = program_level,
        week_strategy = week_strategy
      )
    
    performance_summary <- bind_rows(performance_summary, metrics)
  }
}

# Format and display the summary
if(nrow(performance_summary) > 0) {
  # Create complete performance summary
  performance_display <- performance_summary |>
    select(dataset_id, program_level, week_strategy, .metric, .estimate) |>
    pivot_wider(
      names_from = .metric,
      values_from = .estimate
    ) |>
    arrange(desc(roc_auc))
  
  # Save full performance summary
  saveRDS(performance_summary, 
          file.path(config::get("modelled_dir"), "run", "metrics", "performance_summary.rds"))
  
  # Display overall summary
  cat("\nModel Performance Summary (Top 10, sorted by ROC AUC):\n")
  print(knitr::kable(head(performance_display, 10), digits = 4))
  
  # Compare performance by week strategy using ROC AUC (better for ranking)
  week_strategy_summary <- performance_summary |>
    filter(.metric == "roc_auc") |>  # Using ROC AUC instead of accuracy
    group_by(week_strategy) |>
    summarize(
      avg_roc_auc = mean(.estimate, na.rm = TRUE),
      min_roc_auc = min(.estimate, na.rm = TRUE),
      max_roc_auc = max(.estimate, na.rm = TRUE),
      n_models = n()
    )
  
  cat("\nPerformance Comparison by Week Strategy (ROC AUC):\n")
  print(knitr::kable(week_strategy_summary, digits = 4))
  
  # Visualize strategy comparison with ROC AUC
  ggplot(performance_summary |> filter(.metric == "roc_auc"), 
         aes(x = week_strategy, y = .estimate, fill = week_strategy)) +
    geom_boxplot() +
    labs(
      title = "Model ROC AUC score over tijd",
      subtitle = "Hogere waarden geven betere studentenrangschikkingscapaciteit aan",
      x = "Verloop over tijd",
      y = "ROC AUC"
    ) +
    theme_minimal()
  
  # Check if we have multiple models for same program-level with different strategies
  program_levels_with_all_strategies <- performance_summary |>
    filter(.metric == "roc_auc") |>  # Using ROC AUC for consistent comparison
    group_by(program_level) |>
    summarize(
      n_strategies = n_distinct(week_strategy),
      strategies = paste(sort(unique(week_strategy)), collapse = ", ")
    ) |>
    filter(n_strategies == length(week_strategies))
  
  if(nrow(program_levels_with_all_strategies) > 0) {
    cat("\nProgram-levels with all week strategies available:", 
        nrow(program_levels_with_all_strategies), "\n")
    
    # Compare strategies directly for these program-levels
    direct_comparison <- performance_summary |>
      filter(
        .metric == "roc_auc",  # Using ROC AUC for ranking quality
        program_level %in% program_levels_with_all_strategies$program_level
      ) |>
      select(program_level, week_strategy, roc_auc = .estimate) |>
      pivot_wider(
        names_from = week_strategy,
        values_from = roc_auc
      ) |>
      mutate(
        # Positive values mean the second strategy is better
        early_vs_none = early - none,  # Changed sign to make interpretation consistent
        all_vs_none = all - none,      # Higher values = better ranking
        all_vs_early = all - early
      ) |>
      arrange(desc(all))
    
    cat("\nDirecte vergelijking van week variabelen bij ROC AUC:\n")
    print(knitr::kable(direct_comparison), digits = 2)
    
    # Save this comparison for use in interpretation
    saveRDS(direct_comparison, 
            file.path(config::get("modelled_dir"), "run", "metrics", "week_strategy_comparison.rds"))
  } else {
    cat("\nNo program-levels have models for all week strategies.\n")
  }
} else {
  cat("\nNo model performance metrics available.\n")
}
```

## Volgende stappen

De getrainde modellen en hun prestatiemetrieken zijn nu opgeslagen in de map `data/modelled/run`. In het volgende document zullen we deze modellen interpreteren om de belangrijkste factoren te begrijpen die van invloed zijn op studentuitval.
